A Formalization of Typed Lua

### 2015-08-18 doc/papers/dls15/paper.tex
squeeze the paper into 13 pages after changing text to 10-point

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{url}
\usepackage{color}
\usepackage{multirow}

@@

\begin{figure}[t]
{\footnotesize
\begin{verbatim}
  function new_point(x, y)
    local point = {}
    point.x = x
    point.y = y
    function point:move(dx, dy)
      self.x = self.x + dx
      self.y = self.y + dy
      return self
    end
    return point
  end
\end{verbatim}
}
\caption{Building an object piece-by-piece}
\label{fig:newpoint}
\end{figure}

@@

The rationale for the permissiveness of rule {\sc S-TableUtoC} derives from the fact that reading a non-existing key just yields $\Nil$ instead of being an error in Lua semantics, plus the fact
that the type system guarantees that a table with an
{\em unique} type has no keys with a value that does not
belong to one of its key types, and does not have any
aliases. With these guarantees, a table with type $\{1{:}\String, 2{:}\String,
``x"{:}\Integer, ``y"{:}\Value\}_{unique}$ can be safely recast as a table with type 
$\{\Number{:}\String \cup \Nil, ``x"{:}\Number, ``z"{:}
\Number \cup \Nil\}_{closed}$ as long as it becomes
inaccessible with the original type, which Section~\ref{sec:tables} will show that is the case
for any type judgment where this subtyping rule is used.

@@

In the implementation of Typed Lua, we also use consistent-subtyping to normalize and simplify union types.
For instance, the union type $\Boolean \cup \Any$ simplifies
to $\Any$, because $\Boolean$ is a consistent-subtype
of $\Any$. Another example is the union type
$\Number \cup \Nil \cup 1$, that simplifies to
$\Number \cup \Nil$, because $1$ is a consistent-subtype of $\Number$.

@@

The simplest expression involving tables is the empty
table constructor $\{\}$. The empty table constructor always
has type $\{\}_{unique}$. A more interesting example is $\{ [``x"] = 1, [``y"] = 2 \}$, which has type $\{ ``x":1 , ``y": 2 \}_{unique}$. The subtyping rule for {\em unique} tables lets
us assign this table to a variable with a more general
type such as $\{ ``x":\Number , ``y": \Number \}_{unique}$ or
even $\{ \String:\Number \cup \Nil \}_{unique}$.

@@

Rule {\sc T-Constructor} also checks if the resulting type is {\em well-formed} with the predicate $wf$. Checking for well-formedness avoids ambiguous table types like
$\{1:\Number, \Integer:\String, \Any:\Boolean\}$,
where indexing it with a key of type $\Integer$ could
yield either a number, a string or a boolean at runtime.

@@

There are two ways the type of a {\em unique} and {\em open} tables can change in Typed
Lua. The first way is by assignment to a non-existing
field. Let us go back to the example shown in Figure~\ref{fig:newpoint}. Assuming that the type of the
$x$ and $y$ is $\Number$, we can translate the first three lines
of function {\tt new\_point} to the abstract syntax of Figure~\ref{fig:syntax}:
{\small
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; point: \{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$point[``x"] \; {<}\Number{>} = x;$}\\
& \multicolumn{1}{l}{$point[``y"] \; {<}\Number{>} = y$}
\end{tabular}
\end{center}
}

@@

We could add both fields with a single assignment statement,
as in the following example:
{\small
\begin{center}
	\begin{tabular}{ll}
		\multicolumn{2}{l}{$\mathbf{local} \; point: \{\}_{unique} = \{\} \; \mathbf{in}$}\\
		& \multicolumn{1}{l}{$point[``x"] \; {<}\Number{>},point[``y"] \; {<}\Number{>} = x,y;$}
	\end{tabular}
\end{center}
}
In this case, the first coercion expression would still change the type of $point$ to $\{``x":\Number\}_{unique}$, but the
second would change it to $\{``y":\Number\}_{unique}$.
Merging the two yields $\{``x":\Number, ``y": \Number\}_{unique}$
through the third case of function $merget$ defined in Figure~\ref{fig:funcs}.

@@

Consider the following example, where we make an alias
of a {\em unique} table:
{\small
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{closed} = a \; \mathbf{in}\;\;a[``x"] \; {<}\String{>} = ``foo";$}
\end{tabular}
\end{center}
}

@@

In the example below, we use a coercion expression to
widen the type of a field in a {\em unique}
table, and then add another field, to show that the
table is now {\em open}:
{\small
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{3}{l}{$a[``y"] \; {<}\String{>} = ``bar";$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
& & \multicolumn{2}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{closed}{>} \; a \; \mathbf{in}$}\\
& & & \multicolumn{1}{l}{$a[``z"] \; {<}\Integer{>} = 1$}
\end{tabular}
\end{center}
}

@@

Lua has first-class functions, so another way of aliasing
an {\em unique} and {\em open} references is by closing
over them in a function body. In the environment used for
typing the function body, all {\em unique} and {\em open}
tables are {\em closed} with the auxiliary function $closeall$.
Assignment to a closed-over table can be unsafe, because the {\em reference} to it is aliased as well. In the following
example, assume that $g$ receives a parameter of type
$\{ ``x": \String \cup \Nil\}_{closed}$:
{\small
\begin{center}
	\begin{tabular}{llll}
		\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{open}
			= \{\} \; \mathbf{in}$}\\
		& \multicolumn{3}{l}{$\mathbf{local} \; f:\Void \rightarrow \Void =$}\\
		& & \multicolumn{2}{l}{$\mathbf{fun} \; ():\Void\;\;\;
		 a = \{[``x"]=5\} $}\\
		& \multicolumn{3}{l}{$\mathbf{in} \; \lfloor g({<}\{``x":
			 \String \cup \Nil\}_{closed}{>} \; a) \rfloor_{0}$}
	\end{tabular}
\end{center}
}

@@

The following example typechecks, and uses all the parts of {\sc T-Function}:
{\small
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique}, b:\{\}_{unique} = \{\}, \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Void \rightarrow \Integer \times \Void =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Void$}\\
& & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; \lfloor f(a[``x"]) \rfloor_0$}\\
\end{tabular}
\end{center}
}

@@

\begin{figure*}[t]
{\footnotesize
\label{fig:projtypes}
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-LocalProj}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash el:E, \env_{2}, (x,S)\\
	\env_{3} = \env_{2}[id_{1} \mapsto proj(E, 1), \ldots, id_{n} \mapsto proj(E, n)]\\
	\env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
	n = |\;\overline{id}\;|  
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, (\env_{4}-\{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
\end{array}
\begin{array}{c}
\mylabel{T-ExpListProj}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
	\env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\; n=|\overline{e}|\\
	S = S_1 \sqcup S_2 \;\;\;m = max(|S_1|, |S_2|) \;\;\; x \notin \Pi\\
	\env_{f} = merge(\env_{1}, ..., \env_{n+2})
	\end{array}}
{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x}{*}, \env_{f}, (x,S)}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-IfFilterProjThen}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\ S_t = fonilpt(\Pi(x), i)\\ 
	finil(proj(\Pi(x), i)) = {\bf void} \\
	\env_{1}, \penv[x \mapsto S_t] \vdash s_{1}, \env_{2} 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-IfFilterProjElse}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\
	 S_e = finilpt(\Pi(x), i)\\
	fonil(proj(\Pi(x), i)) = {\bf void} \\
	\env_{1}, \penv[x \mapsto S_e] \vdash s_{2}, \env_{2} 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-IfFilterProjBoth}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\ S_t = fonilpt(\Pi(x), i) \; S_e = finilpt(\Pi(x), i)\\
	\env_{1}, \penv[x \mapsto S_t] \vdash s_{1}, \env_{2} \\
	\env_{1}, \penv[x \mapsto S_e] \vdash s_{2}, \env_{3} \\
	\env_{4} = join(\env_{2}, \env_{3}) 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}\\\\
%\begin{array}{rcll}
%last(F_1{*} \sqcup F_2{*}) & = & F_1{*} & \mathrm{if} \; F_2{*} \subtype F_1{*}\\
%last(F_1{*} \sqcup F_2{*}) & = & F_2{*} & \mathrm{if} \; F_1{*} \subtype F_2{*}\\
%last(F_1{*} \sqcup F_2{*}) & = & (F_1 \cup F_2){*} & \mathrm{otherwise}\\
%last(F{*} \sqcup T \times P) & = & last(\Nil{*} \sqcup P) \\
%last(T \times P \sqcup F{*}) & = & last(P \sqcup \Nil{*}) \\
%last(P \sqcup S) & = & last(P \sqcup last(S)) \\
%last(S \sqcup P) & = & last(last(S) \sqcup P) \\
%last(S_1 \sqcup S_2) & = & last(last(S_1) \sqcup last(S_2))
%\end{array}\\\\
\begin{array}{rcll}
proj(S_1 \sqcup S_2, i) & = & proj(S_1, i) \cup proj(S_2, i) \\
proj(F{*}, i) & = & nil(F) \\
proj(F \times P, 1) & = & F \\
proj(F \times P, i) & = & proj(P, i-1)\\
proj(E{*}, i) & = & nil(E) \\
proj(T \times E, 1) & = & T \\
proj(T \times E, i) & = & proj(E, i-1)
\end{array}\\\\
\begin{array}{rcll}
fonilpt(P_1 \sqcup P_2) & = & P_2 & \mathrm{if} \; fonil(proj(P_1, i)) = {\bf void} \\
fonilpt(P_1 \sqcup P_2) & = & P_1 & \mathrm{if} \; fonil(proj(P_2, i)) = {\bf void} \\
fonilpt(P_1 \sqcup P_2) & = & P_1 \sqcup P_2 & \mathrm{otherwise}\\
fonilpt(P \sqcup S) & = & fonilpt(S) & \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
fonilpt(P \sqcup S) & = & P \sqcup fonilpt(S) & \mathrm{otherwise}\\
fonilpt(S \sqcup P_2) & = & fonilpt(S) & \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
fonilpt(S \sqcup P) & = & fonilpt(S) \sqcup P & \mathrm{otherwise}\\
fonilpt(S_1 \sqcup S_2) & = & fonilpt(S_1) \sqcup fonilpt(S_2)
\end{array}\\\\
\begin{array}{rcll}
fonil(F_1 \cup F_2) & = & fonil(F_1) & \mathrm{if}\;fonil(F_2) = {\bf void}\\
fonil(F_1 \cup F_2) & = & fonil(F_2) & \mathrm{if}\;fonil(F_1) = {\bf void}\\
fonil(F_1 \cup F_2) & = & fonil(F_1) \cup fonil(F_2) & \mathrm{otherwise}\\
%\multicolumn{4}{c}{\mathrm{the\;next\;two\;cases\;apply\;if\;}F
%	\;\mathrm{is\;not\;a\;union}}\\
fonil(F) & = & {\bf void} & \mathrm{if} \; F \subtype \Nil \; \mathrm{and} \; \Nil \subtype F\\
fonil(F) & = & F & \mathrm{otherwise}
\end{array}
\end{array}
\]}
\caption{Rules for Projection Types}
\end{figure*}

@@

As is also shown in Figure~\ref{fig:over}, the idiom for using such
a function is to test the first returned value. We
can translate that code to the abstract syntax of
Figure~\ref{fig:syntax}, leaving placeholders for
the body of the {\bf then} and {\bf else} branches:
{\small
\begin{center}
	\begin{tabular}{ll}
		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(a, b) \, \mathbf{in} \, \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$}
	\end{tabular}
\end{center}
}

@@

Assignment to an l-value with a projection type is forbidden
(there is no rule {\sc T-IdWriteProj}). Allowing assignment is unsound, as it can break the dependency among the components of the union. In the example below,
the projected type of $r$ outside of the {\bf if} statement
is $\Integer \cup \String$, so the assignment seems correct,
but inside $s_1$ the projected type of $r$ would still be
$\Integer$:
{\small
\begin{center}
	\begin{tabular}{ll}
		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(a, b) \, \mathbf{in}$} \\
		&  $r = ``foo" ;\; \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$
	\end{tabular}
\end{center}
}

>>>>

%\usepackage[compact]{titlesec}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{url}
\usepackage{color}
\usepackage{multirow}

\setlength{\textfloatsep}{10pt}

@@

\begin{figure}[t]
{\footnotesize
\begin{verbatim}
  function new_point(x, y)
    local point = {}
    point.x, point.y = x, y
    function point:move(dx, dy)
      self.x, self.y = self.x + dx, self.y + dy
    end
    return point
  end
\end{verbatim}
}
\caption{Building an object piece-by-piece}
\label{fig:newpoint}
\end{figure}

@@

The rationale for the permissiveness of {\sc S-TableUtoC} derives from the fact that reading a non-existing key just yields $\Nil$ instead of being an error in Lua semantics, plus the fact
that the type system guarantees that a table with an
{\em unique} type has no keys with a value that does not
belong to one of its key types, and does not have any
aliases. With these guarantees, a table with type $\{1{:}\String, 2{:}\String,
``x"{:}\Integer, ``y"{:}\Value\}_{unique}$ can be safely recast as a table with type 
$\{\Number{:}\String \cup \Nil, ``x"{:}\Number, ``z"{:}
\Number \cup \Nil\}_{closed}$ as long as it becomes
inaccessible with the original type, which Section~\ref{sec:tables} will show that is the case
for any type judgment where this subtyping rule is used.

@@

The implementation of Typed Lua also uses consistent-subtyping to normalize and simplify union types.
For instance, the union type $\Boolean \cup \Any$ simplifies
to $\Any$, because $\Boolean$ is a consistent-subtype
of $\Any$. Another example is the union type
$\Number \cup \Nil \cup 1$, that simplifies to
$\Number \cup \Nil$, because $1$ is a consistent-subtype of $\Number$.

@@

The simplest expression involving tables is the empty
table constructor $\{\}$. The empty table constructor always
has type $\{\}_{unique}$. A more interesting example is $\{ [``x"] = 1, [``y"] = 2 \}$, which has type $\{ ``x":1 , ``y": 2 \}_{unique}$. The subtyping rule for {\em unique} tables lets
us assign this table to a variable with a more general
type: $\{ ``x":\Number , ``y": \Number \}_{unique}$, or
even $\{ \String:\Number \cup \Nil \}_{unique}$.

@@

Rule {\sc T-Constructor} also checks if the resulting type is {\em well-formed} with the predicate $wf$. Checking for well-formedness avoids ambiguous table types. An example of ambiguous table type is
$\{1:\Number, \Integer:\String, \Any:\Boolean\}$,
where indexing it with a key of type $\Integer$ could
yield either a number, a string or a boolean at runtime.

@@

There are two ways the type of a {\em unique} and {\em open} tables can change in Typed
Lua. The first way is by assignment to a non-existing
field. Let us go back to the example shown in Figure~\ref{fig:newpoint}. Assuming that the type of the
$x$ and $y$ is $\Number$, we can translate the first three lines
of function {\tt new\_point} to the abstract syntax of Figure~\ref{fig:syntax}:
{\footnotesize
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; point: \{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$point[``x"] \; {<}\Number{>} = x;\;
point[``y"] \; {<}\Number{>} = y$}
\end{tabular}
\end{center}
}

@@

We could add both fields with a single assignment statement,
as in the following example:
{\footnotesize
\begin{center}
	\begin{tabular}{ll}
		\multicolumn{2}{l}{$\mathbf{local} \; point: \{\}_{unique} = \{\} \; \mathbf{in}$}\\
		& \multicolumn{1}{l}{$point[``x"] \; {<}\Number{>},point[``y"] \; {<}\Number{>} = x,y;$}
	\end{tabular}
\end{center}
}
In this case, the first coercion expression would still change the type of $point$ to $\{``x":\Number\}_{unique}$, but the
second would change it to $\{``y":\Number\}_{unique}$.
Merging the two yields $\{``x":\Number, ``y": \Number\}_{unique}$
through the third case of function $merget$ defined in Figure~\ref{fig:funcs}.

@@

Consider the following example, where we make an alias
of a {\em unique} table:
{\footnotesize
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{closed} = a \; \mathbf{in}\;\;a[``x"] \; {<}\String{>} = ``foo";$}
\end{tabular}
\end{center}
}

@@

In the example below, we use a coercion expression to
widen the type of a field in a {\em unique}
table, and then add another field, to show that the
table is now {\em open}:
{\footnotesize
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$a[``x"] \; {<}\String{>} = ``foo";\;
a[``y"] \; {<}\String{>} = ``bar";$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
& & \multicolumn{2}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{closed}{>} \; a \; \mathbf{in}$}\\
& & & \multicolumn{1}{l}{$a[``z"] \; {<}\Integer{>} = 1$}
\end{tabular}
\end{center}
}

@@

Lua has first-class functions, so another way of aliasing
an {\em unique} and {\em open} references is by closing
over them in a function body. In the environment used for
typing the function body, all {\em unique} and {\em open}
tables are {\em closed} with the auxiliary function $closeall$.
Assignment to a closed-over table can be unsafe, because the {\em reference} to it is aliased as well. In the following
example, assume that $g$ receives a parameter of type
$\{ ``x": \String \cup \Nil\}_{closed}$:
{\footnotesize
\begin{center}
	\begin{tabular}{llll}
		\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{open}
			= \{\} \; \mathbf{in}$}\\
		& \multicolumn{3}{l}{$\mathbf{local} \; f:\Void \rightarrow \Void =\;\mathbf{fun} \; ():\Void\;\;\;
		 a = \{[``x"]=5\} $}\\
		& & \multicolumn{2}{l}{$\mathbf{in} \; \lfloor g({<}\{``x":
			 \String \cup \Nil\}_{closed}{>} \; a) \rfloor_{0}$}
	\end{tabular}
\end{center}
}

@@

The following example typechecks, and uses all the parts of {\sc T-Function}:
{\footnotesize
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique}, b:\{\}_{unique} = \{\}, \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Void \rightarrow \Integer \times \Void =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Void$}\\
& & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; \lfloor f(a[``x"]) \rfloor_0$}\\
\end{tabular}
\end{center}
}

@@

\begin{figure*}[t]
{\footnotesize
\label{fig:projtypes}
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-LocalProj}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash el:E, \env_{2}, (x,S)\\
	\env_{3} = \env_{2}[id_{1} \mapsto proj(E, 1), \ldots, id_{n} \mapsto proj(E, n)]\\
	\env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
	n = |\;\overline{id}\;|  
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, (\env_{4}-\{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
\end{array}
\begin{array}{c}
\mylabel{T-ExpListProj}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
	\env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\; n=|\overline{e}|\\
	S = S_1 \sqcup S_2 \;\;\;m = max(|S_1|, |S_2|) \;\;\; x \notin \Pi\\
	\env_{f} = merge(\env_{1}, ..., \env_{n+2})
	\end{array}}
{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x}{*}, \env_{f}, (x,S)}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-IfFilterProjThen}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\ S_t = fonilpt(\Pi(x), i)\\ 
	finil(proj(\Pi(x), i)) = {\bf void} \\
	\env_{1}, \penv[x \mapsto S_t] \vdash s_{1}, \env_{2} 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-IfFilterProjElse}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\
	 S_e = finilpt(\Pi(x), i)\\
	fonil(proj(\Pi(x), i)) = {\bf void} \\
	\env_{1}, \penv[x \mapsto S_e] \vdash s_{2}, \env_{2} 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-IfFilterProjBoth}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\ S_t = fonilpt(\Pi(x), i) \; S_e = finilpt(\Pi(x), i)\\
	\env_{1}, \penv[x \mapsto S_t] \vdash s_{1}, \env_{2} \\
	\env_{1}, \penv[x \mapsto S_e] \vdash s_{2}, \env_{3} \\
	\env_{4} = join(\env_{2}, \env_{3}) 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}
\end{array}
\]}
\caption{Rules for Projection Types}
\end{figure*}

\begin{figure}[t]
{\scriptsize
\label{fig:projtypes}
\[
\begin{array}{c}
\begin{array}{rcl}
fonilpt(P_1 \sqcup P_2) & = & P_2 \;\; \mathrm{if} \; fonil(proj(P_1, i)) = {\bf void} \\
fonilpt(P_1 \sqcup P_2) & = & P_1 \;\; \mathrm{if} \; fonil(proj(P_2, i)) = {\bf void} \\
fonilpt(P_1 \sqcup P_2) & = & P_1 \sqcup P_2 \;\; \mathrm{otherwise}\\
fonilpt(P \sqcup S) & = & fonilpt(S) \;\; \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
fonilpt(P \sqcup S) & = & P \sqcup fonilpt(S) \;\; \mathrm{otherwise}\\
fonilpt(S \sqcup P_2) & = & fonilpt(S) \;\; \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
fonilpt(S \sqcup P) & = & fonilpt(S) \sqcup P \;\; \mathrm{otherwise}\\
fonilpt(S_1 \sqcup S_2) & = & fonilpt(S_1) \sqcup fonilpt(S_2)
\end{array}\\\\
\begin{array}{rcl}
fonil(F_1 \cup F_2) & = & fonil(F_1) \;\; \mathrm{if}\;fonil(F_2) = {\bf void}\\
fonil(F_1 \cup F_2) & = & fonil(F_2) \;\; \mathrm{if}\;fonil(F_1) = {\bf void}\\
fonil(F_1 \cup F_2) & = & fonil(F_1) \cup fonil(F_2) \;\; \mathrm{otherwise}\\
%\multicolumn{4}{c}{\mathrm{the\;next\;two\;cases\;apply\;if\;}F
%	\;\mathrm{is\;not\;a\;union}}\\
fonil(F) & = & {\bf void} \;\; \mathrm{if} \; F \subtype \Nil \; \mathrm{and} \; \Nil \subtype F\\
fonil(F) & = & F \;\; \mathrm{otherwise}
\end{array}
\end{array}
\]}
\caption{Auxiliary Functions for Projection Types}
\end{figure}

@@

As is also shown in Figure~\ref{fig:over}, the idiom for using such
a function is to test the first returned value. We
can translate that code to the abstract syntax of
Figure~\ref{fig:syntax}, leaving placeholders for
the body of the {\bf then} and {\bf else} branches:
{\footnotesize
\begin{center}
	\begin{tabular}{ll}
		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(a, b) \, \mathbf{in} \, \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$}
	\end{tabular}
\end{center}
}

@@

Assignment to an l-value with a projection type is forbidden
(there is no rule {\sc T-IdWriteProj}). Allowing assignment is unsound, as it can break the dependency among the components of the union. In the example below,
the projected type of $r$ outside of the {\bf if} statement
is $\Integer \cup \String$, so the assignment seems correct,
but inside $s_1$ the projected type of $r$ would still be
$\Integer$:
{\footnotesize
\begin{center}
	\begin{tabular}{ll}
		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(a, b) \, \mathbf{in}\; r = ``foo" ;\; \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$}
	\end{tabular}
\end{center}
}
